---
title: "Statistical Anlysis of epidemic data and transmission pattern"
author: "Andrea Apolloni & Stéphanie Bougeard "
date: "01/06/2026"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# 0. Summary

Main results

  - Outbreaks appear to occur randomly (Moran I not significative)
  - Clusters analysis using db scan identify easily area where   outbreaks occurred after  vaccination started (cluster per se) and before.  
  - For all the dbscan clusters, outbreaks are less than 20 km from  markets. 
  - In dbscan communes could belong to different clusters and  after checking for time some overlapping is due to local transmission
  - Oriental region appears always to be a cluster per se
  - Not all the clusters are in contact at among themselves. Outliers are the only clusters in contact with all the other
  - At regional level , very few contacts inter regions (1 or 2)  between Marrackeeh and Fes towards  Daraa-Tafilalet
  - At commune level most of the connections (infections) are toward communes in the region of Beni-Mella and Fez Meknes. Probability (frequency) of the contacts are higher that  in the region case

# 1. Introduction


We use several ST-dbscan to identify clusters of propagation for the PPR epidemic in Morocco 2008. 

The two approaches used are:

  - **stdbscan**: a spatio temporal density based clustering algorithm. This algorithm needs 3 inputs : the spatial window, the  temporal window and the minimum number of neighbors. To identify these parameters we first did a sensitivity analysis and select  the "best partition" based on the number of  clustered nodes and the fraction of outliers. To overcome certain points (like small clusters and /or high overlapping  at Admin Units) we  consider other 2 modifications of the algorithm
  
    - **stdbscan_hcpc** after the stdbscan clustering we run Hcpc to aggregate smaller clusters to larger
    

  
In a second step: we used the Wallinga Teunis algorithm , to identify the "Who-Acquire -Infection-From-Whom" matrix. The approach is based on bayesian algorithm , where generation time is considered and the  potential node that could have infected other node is estimated.  The approach produce  for each possible couple of outbreaks a link a with a certain probability. By generating 1000 simulations we count the fraction of links generated by individuals of different cluster. We use this to estimate a contact matrix between  clusters
 
# 2. Packages et fonctions requises

Package rgdal has been discontinued in 2023 , at its place the packages terra et sp are used


```{r, echo = T, warning = F, message = F}
if(!require(spacetime)){install.packages("spacetime")}
if(!require(sp)){install.packages("sp")}
if(!require(xts)){install.packages("xts")}
if(!require(aqp)){install.packages("aqp")}
if(!require(NbClust)){install.packages("NbClust")}
if(!require(factoextra)){install.packages("factoextra")}

library(spacetime)
library(sp)
library(xts)
library(sf)
library(terra)
library(spdep)

library(readxl)
library(stringi)
library(reshape2)
library(nngeo)
library(mapview)



library(aqp)
library(NbClust)
library(factoextra)
library(dplyr)
library(geosphere)
library(purrr)
library(tidyr)
library(outbreaker2)
library(o2ools)
library(gridExtra)
library(dbscan)
library(tmap)
library(tmap.glyphs)
library(cowplot)
library(GGally)
library(patchwork)
library(viridis)
library(forcats)
library(igraph)
source("stdbscan.R")
```


Function *fviz_nbclust* has been changed to take account of the different  packages used and avoid errors

```{r  plotting function, echo=FALSE}
fviz_nbclust <- function (x, FUNcluster = NULL, method = c("silhouette", "wss", 
                                                           "gap_stat"), diss = NULL, k.max = 10, nboot = 100, verbose = interactive(), 
                          barfill = "steelblue", barcolor = "steelblue", linecolor = "steelblue", 
                          print.summary = TRUE, ...) 
{
  set.seed(123)
  if (k.max < 2) 
    stop("k.max must bet > = 2")
  method = match.arg(method)
  if (!inherits(x, c("data.frame", "matrix")) & !("Best.nc" %in% 
                                                  names(x))) 
    stop("x should be an object of class matrix/data.frame or ", 
         "an object created by the function NbClust() [NbClust package].")
  if (inherits(x, "list") & "Best.nc" %in% names(x)) {
    best_nc <- x$Best.nc
    if (any(class(best_nc) == "numeric") ) 
      print(best_nc)
    else if (any(class(best_nc) == "matrix") )
      .viz_NbClust(x, print.summary, barfill, barcolor)
  }
  else if (is.null(FUNcluster)) 
    stop("The argument FUNcluster is required. ", "Possible values are kmeans, pam, hcut, clara, ...")
  else if (!is.function(FUNcluster)) {
    stop("The argument FUNcluster should be a function. ", 
         "Check if you're not overriding the specified function name somewhere.")
  }
  else if (method %in% c("silhouette", "wss")) {
    if (is.data.frame(x)) 
      x <- as.matrix(x)
    if (is.null(diss)) 
      diss <- stats::dist(x)
    v <- rep(0, k.max)
    if (method == "silhouette") {
      for (i in 2:k.max) {
        clust <- FUNcluster(x, i, ...)
        v[i] <- .get_ave_sil_width(diss, clust$cluster)
      }
    }
    else if (method == "wss") {
      for (i in 1:k.max) {
        clust <- FUNcluster(x, i, ...)
        v[i] <- .get_withinSS(diss, clust$cluster)
      }
    }
    df <- data.frame(clusters = as.factor(1:k.max), y = v, 
                     stringsAsFactors = TRUE)
    ylab <- "Total Within Sum of Square"
    if (method == "silhouette") 
      ylab <- "Average silhouette width"
    p <- ggpubr::ggline(df, x = "clusters", y = "y", group = 1, 
                        color = linecolor, ylab = ylab, xlab = "Number of clusters k", 
                        main = "Optimal number of clusters")
    if (method == "silhouette") 
      p <- p + geom_vline(xintercept = which.max(v), linetype = 2, 
                          color = linecolor)
    return(p)
  }
  else if (method == "gap_stat") {
    extra_args <- list(...)
    gap_stat <- cluster::clusGap(x, FUNcluster, K.max = k.max, 
                                 B = nboot, verbose = verbose, ...)
    if (!is.null(extra_args$maxSE)) 
      maxSE <- extra_args$maxSE
    else maxSE <- list(method = "firstSEmax", SE.factor = 1)
    p <- fviz_gap_stat(gap_stat, linecolor = linecolor, 
                       maxSE = maxSE)
    return(p)
  }
}

.viz_NbClust <- function (x, print.summary = TRUE, barfill = "steelblue", 
                          barcolor = "steelblue") 
{
  best_nc <- x$Best.nc
  if (any(class(best_nc) == "numeric") )
    print(best_nc)
  else if (any(class(best_nc) == "matrix") ) {
    best_nc <- as.data.frame(t(best_nc), stringsAsFactors = TRUE)
    best_nc$Number_clusters <- as.factor(best_nc$Number_clusters)
    if (print.summary) {
      ss <- summary(best_nc$Number_clusters)
      cat("Among all indices: \n===================\n")
      for (i in 1:length(ss)) {
        cat("*", ss[i], "proposed ", names(ss)[i], 
            "as the best number of clusters\n")
      }
      cat("\nConclusion\n=========================\n")
      cat("* According to the majority rule, the best number of clusters is ", 
          names(which.max(ss)), ".\n\n")
    }
    df <- data.frame(Number_clusters = names(ss), freq = ss, 
                     stringsAsFactors = TRUE)
    p <- ggpubr::ggbarplot(df, x = "Number_clusters", 
                           y = "freq", fill = barfill, color = barcolor) + 
      labs(x = "Number of clusters k", y = "Frequency among all indices", 
           title = paste0("Optimal number of clusters - k = ", 
                          names(which.max(ss))))
    return(p)
  }
}
# assign them to the factoextra namespace
environment(fviz_nbclust) <- asNamespace("factoextra")
assignInNamespace("fviz_nbclust",fviz_nbclust,"factoextra")
environment(.viz_NbClust) <- asNamespace("factoextra")
assignInNamespace(".viz_NbClust",.viz_NbClust,"factoextra")
```

The function*create_shared_item_heatmap* has been created  to plot the common elements between the different clusters in a shape of a heatmap. The percentage correspond to the fraction of common links  between cluster i and j  divided by the total number of elements in i (sum ver the column ==1)

```{r common elements cluster heatmap}
create_shared_item_heatmap <- function(data, group_col, item_col, title ) {
  group_sym <- rlang::sym(group_col)
  item_sym <- rlang::sym(item_col)
  
  clusters <- sort(unique(data[[group_col]]))
  pairs <- combn(clusters, 2, simplify = TRUE)
  results <- data.frame()
  
  for (n in seq_len(ncol(pairs))) {
    i <- pairs[1, n]
    j <- pairs[2, n]
    
    items_i <- unique(data %>% filter(!!group_sym == i) %>% pull(!!item_sym))
    items_j <- unique(data %>% filter(!!group_sym == j) %>% pull(!!item_sym))
    
    common_elements <- length(intersect(items_i, items_j))
    
    results <- rbind(results, data.frame(i = i, j = j, Common = 100 * common_elements / length(items_i)))
    results <- rbind(results, data.frame(i = j, j = i, Common = 100 * common_elements / length(items_j)))
  }
  
  
  
  ggplot(results, aes(x = i, y = j, fill = Common)) +
    geom_tile(color = "white") +
    geom_text(aes(label = round(Common, 1)), color = "black") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title, x = group_col, y = group_col, fill = "% Shared") +
    theme_minimal()
}
```
Function to avoid the accent and plain write names

```{r avoid accent function}
strip_diacritics <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  stringi::stri_trans_general(x, "Latin-ASCII")
}
```


# 3. Data

PPR is a directly transmitted disease that affects goats and sheep. The disease appeared in Morocco in 2008. Once outbreaks were declared, agents from ONSSA, the Moroccan veterinary services, proceeded to cull the animals in the outbreak. The data was downloaded from the WAHIS website and contains the date of declaration, geographical coordinates of the outbreak, locality, administrative unit (region), number of cases, number of deaths, and initial herd size by species. Data were geo-referenced and check the administrative units of belonging

Data are stored in "epi_2008_geocoded_Nominatim.xlsx"

```{r, echo = T, warning = F, message = F}
# 0. Import du fichier au format 'data.frame' standard
PPR1 <- read_xlsx("../1_Data/epi_2008_geocoded_Nominatim.xlsx")
PPR<-PPR1
PPR$Lat<-as.numeric(as.character(PPR$New_Latitude))
PPR$Lon<-as.numeric(as.character(PPR$New_Longitude))
PPR$Time<-as.Date(PPR$`Start Date`, format = "%Y-%m-%d")


PPR<-PPR[c("Lon","Lat","Time","Cases","Deaths","REGION","PROVINCE","COMMUNE","Location")]
names(PPR)<-c("Lon","Lat","Time","Cases","Deaths","Region","Province","Commune","Location")
PPR<-PPR[complete.cases(PPR), ]
PPR <- PPR %>%
  mutate(across(where(~ is.character(.x) || is.factor(.x)), strip_diacritics))
PPR$LocRegiCommProv<-paste(PPR$Location,PPR$Region,PPR$Province,PPR$Commune,sep="_")
uniquelocations<-sort(unique(PPR$LocRegiCommProv))
uniquelocationsdf<-data.frame("id"=seq(1:length(uniquelocations)),"LocRegiCommProv"=uniquelocations)
PPR<-merge(PPR,uniquelocationsdf,by="LocRegiCommProv")



PPR<-PPR %>% 
   group_by(id) %>% 
   filter(Time == min(Time)) %>% 
   dplyr::slice(1) %>% # takes the first occurrence if there is a tie
   ungroup()


PPR<-PPR%>%
  group_by(across(c(-Cases,-Deaths)))%>%
  summarise_at(c("Cases","Deaths"), sum)%>%
  ungroup()
PPR<-as.data.frame(PPR)
PPR<-PPR[c("Lon","Lat","Time","Cases","Deaths","Region","Province","Commune","Location","id")]
```

The initial database contained `r dim(PPR1)[1]` records. We cleaned them by removing data without information and considering only the first date of appearance in the locality. Thus, we will accumulate cases by species in each household. In the end, the database represents `r dim(PPR)[1]` households in `r length(unique(PPR$Region))` regions, formerly `r length(unique(PPR$Administrative.Division))`, and `r length(unique(PPR$Province))` during the period from `r min(unique(PPR$Time))` to `r max(unique(PPR$Time))`. A total number of cases `r sum(PPR$Cases)` and deaths `r sum(PPR$Deaths)` were recorded.

The data is then transformed into a spatial dataframe.

```{r transformation en Timespatialdataframe,echo=FALSE}
# 1. Mise au format 'sp' des caractéristiques spatiales
PPR.sp <- SpatialPoints(coords = PPR[, 1:2])

# 2. Mise au format 'xts' des caractéristiques temporelles 
PPR.time <- as.POSIXct(PPR$Time, tz = "GMT")

# 3. Mise au format 'spacetime' de l'ensemble des données
PPR.st  <- STIDF(PPR.sp, PPR.time, PPR[,-c(1:3)])
summary(PPR.st)
```



# 4. Graphical representaiton


## 4.1. Saptial plot at differnt time

Below a plot of the cases registered using their location and time. Colors correspond to the number of cases  registered. As we noticed the maximum extension is using the months of July and August
```{r, echo = T, warning = F, message = F}
stplot(PPR.st[, , 1], cuts=20, colorkey = TRUE)
```


We plot the position of the outbreak comapred to  the market positions as provided in the file "../data/markets/Souks_Maroc.xlsx". In the figgure we color differently if markets are considered eitehr rural  uraban and the outbreak positions. Morveor we consider a sort of  market network based on their proximity. We distinguished between movements between Urba markets and from Rural to Urban market. In the first case we considered that a urban market is connected just to another one , while for the second case we considered the 2 colsest Urban market to a rural One. . We notice that just with this assumption the network is connected.
In the figures we also shown the connection between an outbreaks and the closest market (independently of the type)

```{r market and outbreak plotting }
Outbreaks <- PPR %>%
  filter(!is.na(Lon))%>%
  st_as_sf(coords = c("Lon","Lat")) %>%
  st_set_crs(4326)

markets<-read_xlsx("../1_Data/Souks_Maroc.xlsx")
lon_col<-"Longitude"   
lat_col<-"Latitude"



Markets_p <- markets %>%
  filter(!is.na(Longitude))%>%
  st_as_sf(coords = c("Longitude","Latitude")) %>%
  st_set_crs(4326)

marketsu<-markets[markets$AREA_TYPE=="Urban",c("ID","Longitude","Latitude")]
names(marketsu)<-c("ID_Urb","Lon_Urb","Lat_Urb")
marketsr<-markets[markets$AREA_TYPE=="Rural",c("ID","Longitude","Latitude")]
names(marketsr)<-c("ID_Rur","Lon_Rur","Lat_Rur")


Urban_location <- marketsu %>%
  filter(!is.na(Lon_Urb))%>%
  st_as_sf(coords = c("Lon_Urb","Lat_Urb")) %>%
  st_set_crs(4326)
Rural_location <- marketsr %>%
  filter(!is.na(Lon_Rur))%>%
  st_as_sf(coords = c("Lon_Rur","Lat_Rur")) %>%
  st_set_crs(4326)

st_connect(Urban_location, Urban_location,k=2) 

Plot_market_network<-  mapview::mapview(st_connect(Urban_location, Urban_location,k=3),color="red") +
  mapview::mapview(st_connect(Rural_location, Urban_location,k=2),color="blue")+
  mapview::mapview(st_connect(Outbreaks, Markets_p,k=2),color="orange")+  
  mapview::mapview(Rural_location, color = 'green', col.regions = 'green') + 
  mapview::mapview(Urban_location, color = 'black', col.regions = 'black') +
  mapview::mapview(Outbreaks, color = 'maroon', col.regions = 'maroon') 
  
Plot_market_network
```



We calculate the distributions of distances in the above distances  to imrpve the definiton of distance to consider in the  cluster analysis

```{r dstribution distances}




# Distance Outbreak Market ------------------------------------------------

outbreak_market <- merge(PPR[c("id","Lon","Lat")], markets, by=NULL)

outbreak_market$distance <- distHaversine( 
  outbreak_market[, c("Lon","Lat")],
  outbreak_market[, c("Longitude","Latitude")]
)/1000

shortest_distance_Out_Market <- outbreak_market %>% 
  group_by(id) %>% 
  slice_min(order_by = distance, n = 2, with_ties = TRUE,na_rm=TRUE) %>% 
  ungroup()


stats1 <- shortest_distance_Out_Market %>%
  group_by(AREA_TYPE)%>%
  summarize(
    mean   = mean(distance, na.rm = TRUE),
    median = median(distance, na.rm = TRUE)
  )



hout_market<-ggplot(data=shortest_distance_Out_Market,aes(x=distance))+geom_histogram(aes(fill=AREA_TYPE))+geom_vline(data = stats1, aes(xintercept = mean),  color = "#d95f02", linewidth = 1) +
  geom_vline(data = stats1, aes(xintercept = median), color = "#1b9e77", linetype = "dashed", linewidth = 1) +facet_wrap(~AREA_TYPE)+
  labs(title = "Distance Outbreak Souk",
       subtitle = "Solid = mean; Dashed = median",fill="Souk Type",x="Distance (km)",y="Frequency")+ 
  theme_minimal()
hout_market
ggsave("./FiguresForArticle/DistanceOutSouk.png",hout_market,width=15,height=15,units="cm",dpi=330)
# Distance Rural Market and Urban -----------------------------------------


all_locations <- merge(marketsr, marketsu, by=NULL)

all_locations$distance <- distHaversine( 
  all_locations[, c("Lon_Rur","Lat_Rur")],
  all_locations[, c("Lon_Urb","Lat_Urb")]
)/1000

shortest_distance_Rur_Urb <- all_locations %>% 
  group_by(ID_Rur) %>% 
  slice_min(order_by = distance, n = 2, with_ties = TRUE,na_rm=TRUE) %>% 
  ungroup()



stats2 <- shortest_distance_Rur_Urb %>%
  summarize(
    mean   = mean(distance, na.rm = TRUE),
    median = median(distance, na.rm = TRUE)
  )



hrur_urb<-ggplot(data=shortest_distance_Rur_Urb,aes(x=distance))+geom_histogram(fill="green")+geom_vline(data = stats2, aes(xintercept = mean),  color = "#d95f02", linewidth = 1) +
  geom_vline(data = stats2, aes(xintercept = median), color = "#1b9e77", linetype = "dashed", linewidth = 1) +
  labs(title = "Distance Rural Urban",
       subtitle = "Solid = mean; Dashed = median") +
  theme_minimal()


# Distance Urban Urban Market ---------------------------------------------
mat <- distm(
  marketsu[,c("Lon_Urb","Lat_Urb")],
  marketsu[,c("Lon_Urb","Lat_Urb")],
  fun = distHaversine
)

# assign the name to the point in list1 based on shortest distance in the matrix

rownames(mat) <- marketsu$ID_Urb
colnames(mat) <- marketsu$ID_Urb

df <- na.omit(melt(mat))  # reshaping
colnames(df) <- c("Origin", "Destination", "Dist") 
df$Dist<-df$Dist/1000
df<-df[df$Origin !=df$Destination,]

shortest_distance_Urb_Urb <- df %>% 
  group_by(Origin) %>% 
  slice_min(order_by = Dist, n = 2, with_ties = TRUE,na_rm=TRUE) %>% 
  ungroup()
stats3 <- shortest_distance_Urb_Urb  %>%
  summarize(
    mean   = mean(Dist, na.rm = TRUE),
    median = median(Dist, na.rm = TRUE)
  )



hurb_urb<-ggplot(data=shortest_distance_Urb_Urb,aes(x=Dist))+geom_histogram(fill="white",col="black")+geom_vline(data = stats3, aes(xintercept = mean),  color = "#d95f02", linewidth = 1) +
  geom_vline(data = stats3, aes(xintercept = median), color = "#1b9e77", linetype = "dashed", linewidth = 1) +
  labs(title = "Distance Urban Urban",
       subtitle = "Solid = mean; Dashed = median") +
  theme_minimal()


histos<-hrur_urb+hurb_urb
ggsave("./FiguresForArticle/DistanceSouks.png",histos,width=15,height=15,units="cm",dpi=330)
histos
```

# 5. Spatio-temporal Clusters

## 5.1. Data

The data used in this classification are:

- The variable to be explained **y** (each row corresponds to a household),
- its spatial characteristics,
- its temporal characteristics.

Spatial characteristics are generally geographic coordinates (i.e., longitude and latitude). In order to more easily enter the spatial parameters of the classification, it is advisable to convert them to the UTM (Universal Transverse Mercator) metric system by projection. 

Below is an example of data formatted correctly:

```{r, echo = T, warning = F, message = F}


# 1. Mise au format 'date' des caractéristiques temporelles (en jour)

PPR$Time.num   <- as.numeric(PPR$Time)
PPR$Time.recal <- PPR$Time.num - min(PPR$Time.num) 

# 2. Mise au format (UTM) métrique (en km) des coordonnées géographiques (Lon & Lat)
PPR_temp              <- PPR
sp::coordinates(PPR_temp) <- ~ Lon + Lat
sp::proj4string(PPR_temp) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
PPR_temp              <- spTransform(PPR_temp, CRS = "+proj=utm +zone=32 +units=m")
PPR$Lon.UTM           <- sp::coordinates(PPR_temp)[, 1] / 1000
PPR$Lat.UTM           <- sp::coordinates(PPR_temp)[, 2] / 1000

summary(PPR)
```

```{r shapefile}
regions<-unique(PPR$Region)
maroc3<-st_read("../1_Data/Morocco_Admin_Boundaries_2023/New_SHP_Morocco.shp")#mar_admbnda_adm3_hcp_20230925.shp
maroc3 <- maroc3 %>%
  mutate(across(where(~ is.character(.x) || is.factor(.x)), strip_diacritics))
maroc3 <- maroc3 %>%
  filter(REGION %in% regions)

maroc1<- maroc3 %>%
  group_by(ID_REGION) %>% #ADM1_PCODE
  summarise(geometry = st_union(geometry), .groups = "drop")
```
### 5.1.1 Moran I for spatial autocorreamtion

We test if communes with highest prevalence are dispersed or not and in case if clustered or not

```{r Moran I preparing data}
popPR<-read_xlsx("../1_Data/sheep_goat_populations_communes_GLW_v2.xlsx")
popPR <- popPR %>%
  mutate(across(where(~ is.character(.x) || is.factor(.x)), strip_diacritics))
popPR$Tot2010<-popPR$sheep_2010+popPR$goat_2010
PPR_commune<-PPR%>%
  group_by(Region,Province,Commune)%>%
  summarize(Tot_Cases=sum(Cases))
names(PPR_commune)<-c("REGION","PROVINCE","COMMUNE","Tot_Case")
PPR_commune<-merge(PPR_commune,popPR,by=c("COMMUNE","REGION"))
PPR_commune<-PPR_commune%>%
  mutate(Tot_Case=replace_na(Tot_Case,0))%>%
  mutate(Prev=100*Tot_Case/Tot2010)

prevalence_commune<-maroc3%>%
  left_join(PPR_commune,by=c("REGION","PROVINCE","COMMUNE"))%>%
  mutate(Prev=replace_na(Prev,0))

plot_prev<-tm_shape(prevalence_commune) + tm_fill(col="Prev",  palette="Greens") +
              tm_legend(outside=TRUE)

boxplot(prevalence_commune$Prev, horizontal = TRUE)

```

```{r Moran I analysing}
nb <- poly2nb(prevalence_commune, queen=TRUE) # defining neighbours
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
I <- moran(prevalence_commune$Prev, lw, length(nb), Szero(lw))[1]
I
```
```{r Moran I hypothese test}
MC<- moran.mc(prevalence_commune$Prev, lw, nsim=999)
MC
# View results (including p-value)
plot(MC)
```

The value of the Moran'I and its p-value indicate that date are distributed "randomly"

## 5.2. Spatio-temporal classification

### 5.2.1. st-dbscan classification


- Density methods, e.g., **st-dbscan** [Birant, 2007], available on R (https://github.com/Kersauson/ST-DBSCAN) or Python (https://pypi.org/project/st-dbscan/). Their algorithm is based on density calculated using two distance radii, spatial and temporal. The density estimate corresponds to the number of cases in a neighborhood. Euclidean distance is used to calculate neighborhood distances. The advantages of this method are that it is fast and can detect clusters of different shapes (even irregular ones).
 

The st-dbscan method is preferred. It is based on three parameters that influence the number of classes:

- *minpts*: minimum number of observations in a class (= number of neighbors in KNN-type classifications),
- *eps*: minimum radius or distance between two classes (i.e., spatial window),
- *eps2*: minimum radius or distance between two classes (i.e., temporal window).

The lower the values of these parameters, the greater the number of classes obtained. 
If certain observations do not meet the conditions set by the three parameters, they are not classified by the algorithm (they are classified as outliers and assigned to class 0).


### 5.2.2. Number of possible classes

The st-dbscan method provides the assignments (and non-assignments for observations considered outliers) of observations to classes. Consequently, it provides the number of classes. However, optimizing the values of the three parameters is not straightforward. It is advisable to have a preliminary idea of the number of classes. To do this, a large number of indices (detailed in ?NbClust)---calculated from the Euclidean distance matrix calculated over time and metric coordinates---are available.

```{r, echo = T, warning = F, message = F}
res.nbclust <- NbClust(PPR[, c("Time.recal","Lon.UTM","Lat.UTM")], distance = "euclidean", min.nc = 2, max.nc = 20, method = "complete", index = "all")
pfviz<-fviz_nbclust(res.nbclust) + theme_minimal() + ggtitle("NbClust's optimal number of clusters")

ggsave("./FiguresForArticle/NumberofClusters.png",pfviz,width=15,height=15,units="cm",dpi=330)
```

From the plot, the best partition appears to be `r max(res.nbclust$Best.partition)`. Other possibilities are 2, 3 and 10.

### 5.2.3. Optimisation des paramètres


The values of the three parameters can be chosen:


- By considering the *minpts* parameter as the number of neighbors (plus 1) and using the decision aids associated with KNN classification (= K nearest neighbors),

- By conducting a sensitivity study, evaluating different parameter values and studying the number of classes, the size of the classes, the number of unclassified observations, and their ease of interpretation.

+ For the *eps* parameter, values between 20 km and `r round(max(stats2$mean+stats3$mean,stats2$median+stats3$median))` km (in increments of 20 km) are tested. The maximum value being chose based on teh sum median and mean values  of the ditances between makets
+ For the *eps2* parameter, values between 2 days and 20 days (with a step size of 2 days) are tested. We assume that exchanges take place once a week


First, it is possible to correctly evaluate the **minpts** parameter based on a kmeans classification and then by evaluating the optimal value of the number of neighbors.


```{r, echo = T, warning = F, message = F}
pfviz2<-fviz_nbclust(PPR[, c("Time.recal","Lon.UTM","Lat.UTM")], kmeans, method = "wss", k.max = 25) + theme_minimal() + ggtitle("the Elbow Method")
pfviz2
ggsave("./FiguresForArticle/NumberofminPoints.png",pfviz2,width=15,height=15,units="cm",dpi=330)
```

The optimal values for the number of neighbors are between 4 and 9 (**I am not sure **). Therefore, the values for the **minpts** parameter (= minimum number of observations in the class) are between 5 and 10.


Next, the possible values for the three parameters are evaluated in a sensitivity study. The selection criteria are: the number of classes (to be interpreted in light of the results from Section 5.2.2), the number of outliers, and the maximum class size.

```{r, echo = T, warning = F, message = F}
maxdist<-300
minpts.seq             <- c(5:10)
eps.seq                <- seq(from = 20, to = maxdist, by = 20)
eps2.seq               <- seq(from = 2, to = 20, by = 2)
res.stdbscan           <- expand.grid(eps.seq, eps2.seq, minpts.seq)
colnames(res.stdbscan) <- c('Sp.window (km)', 'Temp.window (jr)', 'Min.cases')


for (i in 1 : nrow(res.stdbscan)){
  
  res <- stdbscan(x = PPR$Lon.UTM, y = PPR$Lat.UTM, time = PPR$Time, eps = res.stdbscan[i, 1], eps2 = res.stdbscan[i, 2], minpts = res.stdbscan[i, 3])
  
  if (sum(names(table(res$cluster)) == '0') == 0){
    res.stdbscan$Nb.clusters[i]         <- length(table(res$cluster))
    res.stdbscan$Nb.Outliers[i]         <- table(res$cluster)
    res.stdbscan$Pct.Outliers[i]        <- round(table(res$cluster) / nrow(PPR) * 100, 1)
    res.stdbscan$MaxSize.Cluster[i]     <- max(table(res$cluster))
    res.stdbscan$Pct.MaxSize.Cluster[i] <- round(max(table(res$cluster)) / nrow(PPR) * 100, 1)
  } else{
    res.stdbscan$Nb.clusters[i]         <- length(table(res$cluster)[which(names(table(res$cluster)) != '0')])
    res.stdbscan$Nb.Outliers[i]         <- table(res$cluster)[which(names(table(res$cluster)) == '0')]
    res.stdbscan$Pct.Outliers[i]        <- round(table(res$cluster)[which(names(table(res$cluster)) == '0')] / nrow(PPR) * 100, 1)
    res.stdbscan$MaxSize.Cluster[i]     <- max(table(res$cluster)[which(names(table(res$cluster)) != '0')])
    res.stdbscan$Pct.MaxSize.Cluster[i] <- round(max(table(res$cluster)[which(names(table(res$cluster)) != '0')]) / nrow(PPR) * 100, 1)
  }
}
 
```



The decision rules set are that:

- The maximum class size must not exceed 50% of the sample,
- The number of outliers must be less than 10% of the sample,
- The maximum number of classes must not exceed 10.

```{r, echo = T, warning = F, message = F}
res.stdbscan[, 1:3] <- lapply(res.stdbscan[, 1:3], factor)
res2<-res.stdbscan[res.stdbscan$Pct.MaxSize.Cluster<50 & res.stdbscan$Pct.Outliers<10 & res.stdbscan$Nb.clusters<10, ]

summary(res2)
```
It follows that:

- The optimal spatial window (*eps*) is in the range of 220 to 300 km (260 km is chosen),
- The optimal temporal window (*eps2*) is in the range of 2 days (2 days is chosen),
- The minimum number of cases per class is in the range of 5 or 7 (5 is chosen).




### 5.2.4. Results from stdBscan 


```{r, echo = T, warning = F, message = F}
res <- stdbscan(x = PPR$Lon.UTM, y = PPR$Lat.UTM, time = PPR$Time, eps = 260, eps2 = 2, minpts = 5)
PPRcluster<-PPR
PPRcluster$cluster<-res$cluster
```

The cluster classification identify `r length(res$cluster[res$cluster==0])` outliers (classe 0) and `r length(unique(res$cluster))-1` clusters.


## 5.3. Comparing Different methods


In the following we compare the characterisitcs of the differnt type of clustering approcah

### 5.3.1 Temporality and Region distribution 

Vacciantion has started in the second half of Augusut and the last pick of infections has been recorded 2 weeks later. We check the distribution of outbreaks that could have occurred before or after the vacciantion

```{r checking vaccination impact}

vaccine_time<-as.Date("2008-09-01",format="%Y-%m-%d")
PPRclustervacc<-PPRcluster%>%
  mutate(Occurrence=ifelse(Time>vaccine_time,"After","Before"))

PPRclustervacc<-PPRclustervacc[c("id","Region","Province","cluster","Occurrence")]

PPRclustervacc_long<-PPRclustervacc%>% 
  rename("Cluster_ID"="cluster")%>%
  group_by( Cluster_ID, Occurrence) %>%
  summarise(n_count = n(), .groups = "drop")

plot_vacc<-ggplot(data=PPRclustervacc_long,aes(x=Cluster_ID,y=n_count,group=Occurrence,fill=Occurrence))+geom_col(
  position = "fill")+theme(
    legend.position = "bottom",                 # move legend below
    legend.direction = "horizontal"             # make it horizontal
  ) +
  guides(
    fill = guide_legend(ncol = 3)          # if you also have a colour legend
  )      

PPRclustervacc_Reg<-PPRclustervacc%>% 
  rename("Cluster_ID"="cluster")%>%
  group_by( Cluster_ID, Region) %>%
  summarise(n_count = n(), .groups = "drop")

plot_reg<-ggplot(data=PPRclustervacc_Reg,aes(x=Cluster_ID,y=n_count,group=Region,fill=Region))+geom_col(
  position = "fill")+theme(
    legend.position = "bottom",                 # move legend below
    legend.direction = "horizontal"             # make it horizontal
  ) +
  guides(
    fill = guide_legend(ncol = 3)          # if you also have a colour legend
  )  

pf<-plot_reg+plot_vacc
pf
ggsave("./FiguresForArticle/TemporalRegionClusters.png",pf,width=15,height=15,units="cm",dpi=330)
```

### 5.3.2 Size comparison


### 5.3.2a Distance from market

```{r}
PPRcluster_long<-PPRcluster%>% 
  rename("Cluster_ID"="cluster")
distM_cluster<-merge(PPRcluster_long[c("id","Cluster_ID")],shortest_distance_Out_Market[c("id","distance","AREA_TYPE")],by="id")

p_distM<-ggplot(distM_cluster, aes(x = factor(Cluster_ID), y = distance, fill = factor(Cluster_ID),alpha=as.factor(AREA_TYPE))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
scale_alpha_manual(values = c("Rural" = 1.0,"Urban" = 0.2))+guides(alpha = guide_legend(override.aes = list(fill = "grey")))+
coord_flip() +
  labs(
    x = "Cluster_ID",
    y = "Distance (Km)",
    fill = "Cluster_ID",
    alpha="Souk Type"
  ) +
  theme_minimal()
p_distM
ggsave("./FiguresForArticle/DistanceMarketClusters.png",p_distM,width=15,height=15,units="cm",dpi=330)
```

For all the dbscan clusters, outbreaks are less than 20 km from  markets . 

### 5.3.3 Pairs comparison

The number of clusters identify by the different methods is almost equivalent , with the interface having one cluster more by construction and the hcpc 2 less


```{r plotting pairs, echo = T, warning = F, message = F}

PPRcluster_long_summary<-PPRcluster_long%>%
   select(Cluster_ID,Time,Location,Commune,Province, Region,Cases,Deaths)%>%
  group_by(Cluster_ID) %>%
  summarise(unique_Region = n_distinct(Region),unique_Prov = n_distinct(Province),unique_Com = n_distinct(Commune),unique_Loc = n_distinct(Location),begin=min(Time),end=max(Time),TotCases=sum(Cases),TotDeaths=sum(Deaths))


# Example dataframe
# df <- data.frame(ColorVar = ..., FacetVar = ..., Var1 = ..., Var2 = ..., Var3 = ...)

# Create pairwise plots
plotpairs<-ggpairs(PPRcluster_long,
        columns = c("Time", "Lon", "Lat"),
        aes(color = as.factor(Cluster_ID)),
  upper = list(continuous = "smooth"), # Remove upper triangle
  diag = list(continuous = "densityDiag"), # Remove diagonal
  lower = list(continuous = "points")) 

ggsave("./FiguresForArticle/Corelations.png",plotpairs,width=15,height=15,units="cm",dpi=330)

```
In the above figures the   possible pairs of geographic , and temporal coordinates are presented .   When we consider Longitude and Latitude separately we notice that the db_scan provide clearer temporal paths for all the clusters (except outliers), however when we plot on the Morocco map (Lon Lat coordinates) the  clusters are  difficult to interpret.

```{r plotting pairs 2, echo = T, warning = F, message = F}




# Example dataframe
# df <- data.frame(ColorVar = ..., FacetVar = ..., Var1 = ..., Var2 = ..., Var3 = ...)

# Create pairwise plots
ggpairs(PPRcluster_long,
        columns = c("Lon", "Lat"),
        aes(color = as.factor(Cluster_ID)),
  upper = list(continuous = "points"), # Remove upper triangle
  diag = list(continuous = "densityDiag"), # Remove diagonal
  lower = list(continuous = "smooth")) 



```

### 5.3.4 Cluster Sizes by Admin Units


```{r resuming  cluster}
PPRcluster_long_summary<-PPRcluster_long%>%
   select(Cluster_ID,Time,Location,Commune,Province, Region,Cases,Deaths)%>%
  group_by(Cluster_ID) %>%
  summarise(unique_Region = n_distinct(Region),unique_Prov = n_distinct(Province),unique_Com = n_distinct(Commune),unique_Loc = n_distinct(Location),begin=min(Time),end=max(Time),TotCases=sum(Cases),TotDeaths=sum(Deaths))
PPRcluster_long_summary
```

```{r summary by cluster}
PPRcluster_long_summary2<-PPRcluster_long_summary%>%
     pivot_longer(cols=c(unique_Loc,unique_Com,unique_Prov,unique_Region),names_to="Admin_Unit",values_to="Unique_Values")

plot_summary2<-ggplot(data=PPRcluster_long_summary2,aes(x=Cluster_ID,y=Unique_Values))+geom_line()+geom_point()+facet_wrap(.~Admin_Unit,scales="free_y")
plot_summary2
ggsave("./FiguresForArticle/SummaryAdminClusters.png",plot_summary2,width=15,height=15,units="cm",dpi=330)
```
### 5.3.5 Cluster Activity




```{r time distribution,echo=FALSE}
pint<-ggplot(PPRcluster_long_summary,aes(
    x = begin, xend = end,   # beginning & end
    y = Cluster_ID,  yend = Cluster_ID,color=as.factor(Cluster_ID)   # keep y fixed
  )) +geom_segment()+labs(color="Cluster_ID")
pint
ggsave("./FiguresForArticle/ActivityTimeClusters.png",pint,width=15,height=15,units="cm",dpi=330)
```
Dbscan clusters identify cluster well identified temporally. In all these clusters outliers (Cluster 0) are present all along the epidemics and should act as bridges between clusters.  We notice moreover that  the first 2  clusters appear also at the same time and evolve in parallel. Similar for all the other clusters they are in parallel 

### 5.3.6 Location distance distribution


```{r distance distribution by cluster,echo=FALSE}
distances <- PPRcluster_long %>%
  group_by(Cluster_ID) %>%
  group_modify(~ {
    coords <- as.matrix(select(.x, Lon, Lat))
    # Compute pairwise distances (in meters)
    d <-distm(coords, fun = distHaversine)/1000
    # Convert upper triangle to tidy form
    as_tibble(as.data.frame(as.table(d))) %>%
      rename(from = Var1, to = Var2, distance = Freq) %>%
      mutate(from = as.integer(from),
             to   = as.integer(to)) %>%   # convert factors → integers
      filter(from < to) 
  })

pdist<-ggplot(data=distances,aes(x=distance,y=as.factor(Cluster_ID),group=as.factor(Cluster_ID), fill=as.factor(Cluster_ID)))+geom_boxplot(median.color="black")+labs(fill="Cluster_ID")
pdist
ggsave("./FiguresForArticle/DistanceLocationClusters.png",pdist,width=15,height=15,units="cm",dpi=330)
```

In the Dbscan cluster the outliers (cluster 0) occur without pattern they are randomly distributed among all territory and so widely spread, also outbreaks in this cluster occur all along the epidemics  . Cluster 1 and Cluster 2 are characterised by a wider distance distribution among outbreaks. The distance distribution could indicate long distance due  animal commercial movement. Cluster 3 (Cluster (2 for hcpc))  could correspond to first introduction area, when the disease has not yet detected and animals have been moved towards markets, while cluster 4 and 5 correspond to the final parts Due their shortest interval and distances could correspond to very local infections  (?). In the interface case , Cluster 6 has the same distance distribution similar to cluster 1 and 2. 



### 5.3.7 Inter-outbreak  interval distribution

```{r times between outbreaks per cluster, echo=FALSE}
PPRcluster_long2<-PPRcluster_long%>%
  arrange(Cluster_ID, Time) %>%         # sort by group and time
  group_by(Cluster_ID) %>%
  arrange(Time) %>%
  mutate(time_diff = Time - lag(Time,default=min(Time)))

global_interval<-unique(PPRcluster[c("Lon","Lat","Time","id")] )%>%
  arrange(Time) %>%         # sort by group and time
  mutate(time_diff = Time - lag(Time,default=min(Time)))
global_interval$Cluster_ID<-max(PPRcluster_long$Cluster_ID)+1



distrinterval<-rbind(PPRcluster_long2[c("Lon","Lat","Time","id","Cluster_ID","time_diff")], global_interval[c("Lon","Lat","Time","id","Cluster_ID","time_diff")])
distrinterval<-distrinterval[complete.cases(distrinterval), ]

plotdistrinterval<-ggplot(distrinterval,aes(x=time_diff,group=as.factor(Cluster_ID),fill=as.factor(Cluster_ID)))+labs(fill="Cluster_ID")+geom_boxplot()
plotdistrinterval
```

```{r summary times between outbreaks per cluster, echo=FALSE}

distrintervalsummary<-distrinterval%>%
  group_by(Cluster_ID)%>%
  summarise(Median=median(time_diff,na.rm=TRUE),
    q025   = quantile(time_diff, 0.025,na.rm=TRUE),
    q975   = quantile(time_diff, 0.975,na.rm=TRUE))
distrintervalsummary
```
In the above figure we plot the time interval between 2 declaration in each cluster (1-5)  , in the outliers (0) and in general without taking consideration of cluster(cluster 8 ) correspond to all the declaration without taking account the cluster by method of detection. In general the lag of time of declaration is less than 5 days However this changes depending on cluster and method used .


### 5.3.8 PPR impact by Cluster

```{r number of cases and deaths in cluster}
cases_epi<-PPRcluster_long[c("Cluster_ID","Cases","Deaths")]
cases_epi<-cases_epi%>%
    pivot_longer(cols=c(Cases,Deaths))

p_epi<-ggplot(data=cases_epi,aes(x=value,y=as.factor(Cluster_ID),group=as.factor(Cluster_ID), fill=as.factor(Cluster_ID),color=as.factor(Cluster_ID)))+geom_boxplot(median.color = "black")
p_epi
```

In the below plot we color the commune  based if they belong to a cluster or multiple clusters. As we can see the optic clustring doesn't present overlap.



### 5.3.10 Heatmaps common elements

We plot then the common elements between 2 clusters in term of commune , province and region where outbreaks of differnt clusters have been collected

```{r creating heatmaps }

listPlots<-list()


for(m in unique(PPRcluster_long$Cluster_Method)){
  PPRcluster_longM<-PPRcluster_long[PPRcluster_long$Cluster_Method==m,]
  
  
common_shared<-create_shared_item_heatmap(PPRcluster_longM, "Cluster_ID", "Commune", title = "Commune")
province_shared<-create_shared_item_heatmap(PPRcluster_longM, "Cluster_ID", "Province", title = "Province")
region_shared<-create_shared_item_heatmap(PPRcluster_longM, "Cluster_ID", "Region", title = "Region")
shared_Admins<-plot_grid(region_shared, province_shared,common_shared, ncol = 3)
listPlots[[m]]<-shared_Admins
}





shared_Admins_method<-plot_grid(listPlots[["dbscan"]],listPlots[["dbscan_hcpc"]],listPlots[["dbscan_interface"]],listPlots[["optics" ]],nrow = 4,labels=unique(PPRcluster_long$Cluster_Method))
shared_Admins_method
```
Without enetring into the details , we notice that the for the optics case there is no overalp a tomune levels between different , outliers included, strengthening the fact that this clustering approach favours the local interactions. For the other methods we notice that this not the same 

# 6.  Understanding who infects who

Based on the declaration dates and cluster distribution , we tray to infer  the transmission pattern. We use  the pacakge outbreaker2 

## 6.1 Preparing the data

We use the outbreaker2 function to estimate who infects who. Using a gamma distribution and calinratiing from the data , it runs several simulations and in ech infer who has infected who . We then associate teh clsute and characcteristics of the foyers

```{r using outbreaker,cahe=TRUE}


# --- your inputs ---
# df: one row per case with columns id, date (numeric or Date), cluster (factor/int)
# w_dens: generation time density vector (e.g. dgamma(1:30, shape=2, scale=3))



cluster_id<-unique(PPRcluster[c("id","Location","Region","Commune","Province","cluster")])
names(cluster_id)<-c("id","Location","Region","Commune","Province","dbscan")
cluster_id<-data.frame(lapply(cluster_id, as.character), stringsAsFactors = FALSE)

# 1) Build data and run outbreaker2
mean_elapsed<-mean(diff(sort(unique(PPRcluster$Time.recal))))
sd_elapsed<-sd(diff(sort(unique(PPRcluster$Time.recal))))
w <-unname(table(diff(sort(PPRcluster$Time.recal)))[1:8])/sum(table(diff(sort(PPRcluster$Time.recal)))[1:8]) # empirical
#w<-dgamma(1:15, shape = 2, scale = 2) #original
final_gen<-c()
for(c in unique(sort(PPRcluster$cluster))){
   
     PPRC<-PPRcluster[PPRcluster$cluster==c,]
    res<-diff(sort(PPRC$Time.recal))
     final_gen<-c(final_gen,res)
 
 
}
wc<-unname(table(final_gen))/length(final_gen)#empirical using cluster
PPRcluster_ord<-PPRcluster%>%
  select(id,Time.recal)%>%
  arrange(Time.recal)

dat <- outbreaker_data(
  dates = PPRcluster_ord$Time.recal,
  ids   = PPRcluster_ord$id,
  w_dens = w
)

set.seed(123)
fit <- outbreaker(data = dat, config = create_config(n_iter = 2e4,sample_every = 20,init_tree ="star"))

# 2) Extract many posterior transmission trees (edges i -> j per iteration)
trees <- get_trees(fit, t_inf = FALSE)  # returns a list of data.frames with columns 'from','to'


```

```{r}
plot(fit,type = "t_inf", burnin = 2000)
```

```{r}
net1<-summary(fit)$tree
write.csv(net1,"../Cluster_Analysis/FiguresForArticle/net_empirical_v110626.csv")
unique(net1$generations)
net1<-net1[c("from","to","support")]
net1<-net1%>%
  drop_na()

net<-graph_from_data_frame(net1,directed=TRUE)
E(net)$weight<-net1$support
l <- layout_with_kk(net)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
plot(net,rescale=F,layout=l)
```


```{r associating to and from,cahe=TRUE}
# 3) Bind all trees, add clusters for from/to, and count

edge_samples<-data.frame()


for(m in seq(1:length(trees))){
  treem<-trees[[m]]
  edge_samples<-rbind(edge_samples,treem)
}





edge_samples<-edge_samples[complete.cases(edge_samples),]

edge_counts <- edge_samples %>%
  group_by(from, to) %>%
  summarise(n = n()/length(trees), .groups = "drop")


edge_counts <- edge_counts %>%
  left_join(cluster_id %>% select(id,Region_from=Region,Province_from=Province,Commune_from=Commune,dbscan_from =dbscan),by = c("from" = "id"),relationship = "many-to-many") %>%
  left_join(cluster_id %>% select(id,Region_to=Region,Province_to=Province,Commune_to=Commune,dbscan_to =dbscan),by = c("to" = "id"),relationship = "many-to-many") %>%
  rename(from = from,to = to)

write.csv(edge_counts,"../Cluster_Analysis/FiguresForArticle/edge_counts_empirical_v110626.csv")

netdata2<-edge_counts[c("from","to","n")]
netdata2<-netdata2%>%
  filter(n>0.1)
net2<-graph_from_data_frame(netdata2,directed=TRUE)
E(net2)$weight<-netdata2$n
```

## 6.2 Contact matrices per cluster

From the the trees we count for each method of clustering the average number of contacts that  an individual in group i made in group j (infections), and the average fresuaency (prob)

```{r estimating contact matrices per cluster}
contact_group_stats <- function(data,group_from,group_to,from,to,strength) {
  
  
  # Tidy-eval with {{}} lets you pass column names unquoted
  per_ind <- data %>%
    group_by({{ group_from }}, {{ group_to }}, {{ from }}) %>%
    summarise(
      n_partners = n_distinct({{ to }}),                       # distinct partners in target group
      avg_strength_per_partner = mean({{ strength }}, na.rm = TRUE), # mean strength across partners
      .groups = "drop"
    )
  
  by_group_pair <- per_ind %>%
    group_by({{ group_from }}, {{ group_to }}) %>%
    summarise(
      avg_contacts = mean(n_partners),                         # average number of contacts per individual
      avg_prob = mean(avg_strength_per_partner, na.rm = TRUE),
      .groups = "drop"
    )
  
  names(by_group_pair)<-c("Cluster_from","Cluster_to","avg_contacts","avg_prob")
  
  return(by_group_pair)
}

dbscanM<-contact_group_stats(edge_counts,dbscan_from,dbscan_to,from,to,n)




probability_estimations<-dbscanM



```



```{r plot heatmap per cluster}

mean_contacts_heatmap<-ggplot(probability_estimations,aes(x = Cluster_from, y = Cluster_to, fill = avg_contacts)) +
  geom_tile()+ scale_fill_gradient(low="white", high="red")+ labs(title = "Average Number of contacts")+theme(
    legend.position = "bottom",                 # move legend below
    legend.direction = "horizontal"             # make it horizontal
  ) +
  guides(
    fill = guide_legend(ncol = 3)          # if you also have a colour legend
  )



mean_p_heatmap<-ggplot(probability_estimations,aes(x = Cluster_from, y = Cluster_to, fill = avg_prob)) +
  geom_tile()+ scale_fill_gradient(low="white", high="red")+ labs(title = "Average Frequency")+theme(
    legend.position = "bottom",                 # move legend below
    legend.direction = "horizontal"             # make it horizontal
  ) +
  guides(
    fill = guide_legend(ncol = 3)          # if you also have a colour legend
  )

matrix_clusters<-mean_contacts_heatmap+mean_p_heatmap

matrix_clusters
ggsave("../3_Outputs/Contact_Cluster_empirical_v110626.png",matrix_clusters)
```

Not all cluster interact among themselves. 
In general, all clusters infect nodes in the outleirs. In the dbscan clusters type , mebers of cluster 1 interact mostly among themselves , although the frequency is low. For optics scan we notice that outside clsuters interaction are low, but freauncy of contacts is almost flat

## 6.3 Contact matrices per region

In asimilar way we estimate the contact matrix par region in this case instead of aggregating  by cluster  we aggregate by region independently of the cluster

```{r estimating probability per Commune}

edge_counts<-read.csv("../3_Outputs/edge_counts_empirical_v110626.csv")

contact_commune_region_stats <- function(data,from, to,commune_from, commune_to,region_from, region_to,strength) {
  
  
  # --- 1) Per-individual metrics by commune and region
  per_ind <- data %>%
    group_by(
      {{ commune_from }}, {{ commune_to }},
      {{ region_from }},  {{ region_to }},
      {{ from }}
    ) %>%
    summarise(
      n_partners = n_distinct({{ to }}),                      # distinct partners
      avg_strength_per_partner = mean({{ strength }}, na.rm = TRUE),
      .groups = "drop"
    )
  
  # --- 2) Commune → Commune aggregation
  by_commune_pair <- per_ind %>%
    group_by({{ region_from }},  {{ region_to }},{{ commune_from }}, {{ commune_to }}) %>%
    summarise(
      avg_contacts = mean(n_partners),
      avg_prob = mean(avg_strength_per_partner, na.rm = TRUE),
      .groups = "drop"
    ) 
  
  names(by_commune_pair) <- c(
    "Region_from","Region_to","Commune_from", "Commune_to",
    "avg_contacts", "avg_prob"
  )
  
  # --- 3) Region → Region aggregation
  by_region_pair <- per_ind %>%
    group_by({{ region_from }}, {{ region_to }}) %>%
    summarise(
      avg_contacts = mean(n_partners),
      avg_prob = mean(avg_strength_per_partner, na.rm = TRUE),
      .groups = "drop"
    ) 
  
  names(by_region_pair) <- c(
    "Region_from", "Region_to",
    "avg_contacts", "avg_prob"
  )
  
  # Output both levels
  return(list(
    by_commune_pair  = by_commune_pair,
    by_region_pair   = by_region_pair
  ))
}
results_matrixgeo<-contact_commune_region_stats(edge_counts,from, to,Commune_from, Commune_to,Region_from, Region_to,n) 
Commune_matrix<-results_matrixgeo[["by_commune_pair"]]
Region_matrix<-results_matrixgeo[["by_region_pair"]]

```

```{r function for plotting heatmap by Region}

plot_heatmap_regions_only <- function(
    data,
    region_from, region_to,
    value,                         # e.g., avg_contacts or avg_prob
    title       = "Contacts between communes",
    fill_label  = "Average contacts",
    palette     = "viridis",       # viridis | magma | plasma | inferno
    na_fill     = "grey90",
    expand_grid = TRUE,            # completes the matrix with NA for missing pairs
    line_col    = "white",
    line_size   = 0.75
) {
  # Tidy-eval
  rf  <- enquo(region_from)
  rt  <- enquo(region_to)
  val <- enquo(value)
  
  # Keep only necessary columns with standard names
  df <- data %>%
    select(
      Region_from  = !!rf,
      Region_to    = !!rt,
      value        = !!val
    )
  
  # Optionally complete the full matrix (all from-communes × to-communes)
  if (isTRUE(expand_grid)) {
    rows_key <- df %>% distinct(Region_from)
    cols_key <- df %>% distinct(Region_to)
    
    df <- tidyr::expand_grid(rows_key, cols_key) %>%
      left_join(df, by = c("Region_from",  "Region_to"))
  }
  
  # Build ordered keys: by Region then Commune (alphabetical)
  rows_key <- df %>%
    distinct(Region_from) %>%
    arrange(Region_from) %>%
    mutate(row_id = row_number())
  
  cols_key <- df %>%
    distinct(Region_to) %>%
    arrange(Region_to) %>%
    mutate(col_id = row_number())
  
  # Attach numeric indices to the main data
  df <- df %>%
    left_join(rows_key, by = c("Region_from")) %>%
    left_join(cols_key, by = c("Region_to")) %>%
    filter(!is.na(row_id), !is.na(col_id))
  
  # Compute region block boundaries and midpoints (for labels)
  row_blocks <- rows_key %>%
    group_by(Region_from) %>%
    summarise(
      start = min(row_id),
      end   = max(row_id),
      mid   = (start + end) / 2,
      .groups = "drop"
    ) %>%
    arrange(start)
  
  col_blocks <- cols_key %>%
    group_by(Region_to) %>%
    summarise(
      start = min(col_id),
      end   = max(col_id),
      mid   = (start + end) / 2,
      .groups = "drop"
    ) %>%
    arrange(start)
  
  # Lines between region blocks sit at "end + 0.5", excluding the final edge
  hlines <- row_blocks$end + 0.5
  hlines <- hlines[hlines < max(rows_key$row_id) + 0.5]
  
  vlines <- col_blocks$end + 0.5
  vlines <- vlines[vlines < max(cols_key$col_id) + 0.5]
  
  # Palette
  scale_fill <- switch(
    palette,
    viridis = scale_fill_viridis_c(option = "viridis", na.value = na_fill),
    magma   = scale_fill_viridis_c(option = "magma",   na.value = na_fill),
    plasma  = scale_fill_viridis_c(option = "plasma",  na.value = na_fill),
    inferno = scale_fill_viridis_c(option = "inferno", na.value = na_fill),
    scale_fill_viridis_c(option = "viridis", na.value = na_fill)
  )
  
  # Plot using numeric indices so we can place region labels at midpoints
  ggplot(df, aes(x = col_id, y = row_id, fill = value)) +
    geom_tile(color = "white", linewidth = 0.15) +
    # Region block lines
    geom_hline(yintercept = hlines, color = line_col, linewidth = line_size) +
    geom_vline(xintercept = vlines, color = line_col, linewidth = line_size) +
    # Axes show ONLY region names (at midpoints)
    scale_y_continuous(
      breaks = row_blocks$mid,
      labels = row_blocks$Region_from,
      expand = c(0, 0)
    ) +
    scale_x_continuous(
      breaks = col_blocks$mid,
      labels = col_blocks$Region_to,
      expand = c(0, 0),
      position = "top"
    ) +
    coord_fixed() +
    scale_fill +coord_flip()+
    labs(
      x = "Region (to)",
      y = "Region (from)",
      fill = fill_label
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.ticks = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5),legend.position = "bottom",legend.direction = "horizontal"
    )
}
```

```{r}



Region_prob_heatmap<-plot_heatmap_regions_only(
  Region_matrix,
  Region_from, Region_to,
  avg_prob,                       # e.g., avg_contacts or avg_prob
  title = "Frequency Contacts between Regions",
  fill_label = "Average Frequency",
  palette = c("viridis", "magma", "plasma", "inferno")[1],
  na_fill = "grey90",
  expand_grid = TRUE           # fill missing (from,to) pairs with NA for a full matrix
)
Region_contacts_heatmap<-plot_heatmap_regions_only(
  Region_matrix,
  Region_from, Region_to,
  avg_contacts,                       # e.g., avg_contacts or avg_prob
  title = "Contacts between Region",
  fill_label = "Average Contacts",
  palette = c("viridis", "magma", "plasma", "inferno")[1],
  na_fill = "grey90",
  expand_grid = TRUE           # fill missing (from,to) pairs with NA for a full matrix
)
Region_contacts_heatmap
ggsave("../3_Outputs/Contact_Region_empirical_v110626.png",Region_contacts_heatmap, width = 10, height = 10, dpi = 300)
```
```{r region frequency}
Region_prob_heatmap
```

Interactions inter-regions are very few , and unfrequent   except  between Marrackeeh and Fes Towrds  Daraa-Tafilalet

## 6.4 Contact matrices per Commune

In asimilar way we estimate the contact matrix par commune in this case instead of aggregating  by cluster  we aggregate by commune independently of the cluster. The number of commune is large so difficult to visulize
However, we notice that most of the connections (infections) are toward communes in the region of Beni-Mella and Fez Meknes. Probability (frequency) of the contacts are higher that  in the region case

```{r function for plotting heatmap by commune}

plot_commune_heatmap_regions_only <- function(
    data,
    commune_from, commune_to,
    region_from, region_to,
    value,                         # e.g., avg_contacts or avg_prob
    title       = "Contacts between communes",
    fill_label  = "Average contacts",
    palette     = "viridis",       # viridis | magma | plasma | inferno
    na_fill     = "grey90",
    expand_grid = TRUE,            # completes the matrix with NA for missing pairs
    line_col    = "white",
    line_size   = 0.75
) {
  # Tidy-eval
  cf  <- enquo(commune_from)
  ct  <- enquo(commune_to)
  rf  <- enquo(region_from)
  rt  <- enquo(region_to)
  val <- enquo(value)
  
  # Keep only necessary columns with standard names
  df <- data %>%
    select(
      Commune_from = !!cf,
      Commune_to   = !!ct,
      Region_from  = !!rf,
      Region_to    = !!rt,
      value        = !!val
    )
  
  # Optionally complete the full matrix (all from-communes × to-communes)
  if (isTRUE(expand_grid)) {
    rows_key <- df %>% distinct(Region_from, Commune_from)
    cols_key <- df %>% distinct(Region_to,   Commune_to)
    
    df <- tidyr::expand_grid(rows_key, cols_key) %>%
      left_join(df, by = c("Region_from", "Commune_from", "Region_to", "Commune_to"))
  }
  
  # Build ordered keys: by Region then Commune (alphabetical)
  rows_key <- df %>%
    distinct(Region_from, Commune_from) %>%
    arrange(Region_from, Commune_from) %>%
    mutate(row_id = row_number())
  
  cols_key <- df %>%
    distinct(Region_to, Commune_to) %>%
    arrange(Region_to, Commune_to) %>%
    mutate(col_id = row_number())
  
  # Attach numeric indices to the main data
  df <- df %>%
    left_join(rows_key, by = c("Region_from", "Commune_from")) %>%
    left_join(cols_key, by = c("Region_to", "Commune_to")) %>%
    filter(!is.na(row_id), !is.na(col_id))
  
  # Compute region block boundaries and midpoints (for labels)
  row_blocks <- rows_key %>%
    group_by(Region_from) %>%
    summarise(
      start = min(row_id),
      end   = max(row_id),
      mid   = (start + end) / 2,
      .groups = "drop"
    ) %>%
    arrange(start)
  
  col_blocks <- cols_key %>%
    group_by(Region_to) %>%
    summarise(
      start = min(col_id),
      end   = max(col_id),
      mid   = (start + end) / 2,
      .groups = "drop"
    ) %>%
    arrange(start)
  
  # Lines between region blocks sit at "end + 0.5", excluding the final edge
  hlines <- row_blocks$end + 0.5
  hlines <- hlines[hlines < max(rows_key$row_id) + 0.5]
  
  vlines <- col_blocks$end + 0.5
  vlines <- vlines[vlines < max(cols_key$col_id) + 0.5]
  
  # Palette
  scale_fill <- switch(
    palette,
    viridis = scale_fill_viridis_c(option = "viridis", na.value = na_fill),
    magma   = scale_fill_viridis_c(option = "magma",   na.value = na_fill),
    plasma  = scale_fill_viridis_c(option = "plasma",  na.value = na_fill),
    inferno = scale_fill_viridis_c(option = "inferno", na.value = na_fill),
    scale_fill_viridis_c(option = "viridis", na.value = na_fill)
  )
  
  # Plot using numeric indices so we can place region labels at midpoints
  ggplot(df, aes(x = col_id, y = row_id, fill = value)) +
    geom_tile(color = "white", linewidth = 0.15) +
    # Region block lines
    geom_hline(yintercept = hlines, color = line_col, linewidth = line_size) +
    geom_vline(xintercept = vlines, color = line_col, linewidth = line_size) +
    # Axes show ONLY region names (at midpoints)
    scale_y_continuous(
      breaks = row_blocks$mid,
      labels = row_blocks$Region_from,
      expand = c(0, 0)
    ) +
    scale_x_continuous(
      breaks = col_blocks$mid,
      labels = col_blocks$Region_to,
      expand = c(0, 0),
      position = "top"
    ) +
    coord_fixed() +
    scale_fill +coord_flip()+
    labs(
      x = "Region (to)",
      y = "Region (from)",
      fill = fill_label
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.ticks = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5),legend.position = "bottom",,
    legend.direction = "horizontal"
    )
}
```

```{r}



Commune_prob_heatmap<-plot_commune_heatmap_regions_only(
  Commune_matrix,
  Commune_from, Commune_to,
  Region_from, Region_to,
  avg_prob,                       # e.g., avg_contacts or avg_prob
  title = "Frequency Contacts between communes",
  fill_label = "Average Frequency",
  palette = c("viridis", "magma", "plasma", "inferno")[1],
  na_fill = "grey90",
  expand_grid = TRUE           # fill missing (from,to) pairs with NA for a full matrix
)
Commune_contacts_heatmap<-plot_commune_heatmap_regions_only(
  Commune_matrix,
  Commune_from, Commune_to,
  Region_from, Region_to,
  avg_contacts,                       # e.g., avg_contacts or avg_prob
  title = "Contacts between communes",
  fill_label = "Average Contacts",
  palette = c("viridis", "magma", "plasma", "inferno")[1],
  na_fill = "grey90",
  expand_grid = TRUE           # fill missing (from,to) pairs with NA for a full matrix
)
Commune_contacts_heatmap
ggsave("../3_Outputs/Contact_Commune_empirical_v110626.png",Commune_contacts_heatmap)
```

```{r commune frequency}
Commune_prob_heatmap
```

# 7. Saving data

```{r save data}
#PPRcluster<-PPRcluster%>%
#  rename(dbscan=cluster)

#write.csv(PPRcluster,"Locations_Cluster_v2026.csv")
#write.csv(probability_estimations,"Commune_Contact_matrix_original_v110626.csv")
#write.csv(edge_counts,"Transmission_network_v2026.csv")
#write.csv(Commune_matrix,"Commune_Contact_matrix_original_v110626.csv")
```






# 8. References

- https://cran.r-project.org/web/views/SpatioTemporal.html (all the related R packages)
- https://spacetimewithr.org/Spatio-Temporal%20Statistics%20with%20R.pdf (Spatio-Temporal Statistics with R; free book)
- https://informatique-mia.inrae.fr/reseau-resste/sites/default/files/2020-09/analyzing-spatio-temporal.pdf (article JSFdS)
- https://informatique-mia.inrae.fr/reseau-resste/spatio-temporal-geostatistics (tutoriels du groupe français RESSTE)
- https://link.springer.com/content/pdf/10.1007/s10462-019-09736-1.pdf (article de revue sur les méthodes de classification spatio-temporelles)



